---
title: "q3"
author: "Silpa"
date: "2025-01-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(biostat3)
library(knitr)
library(survival)
library(collett)
```

## (a) Plots of survival

```{r}
# Prepare the data for survival analysis
melanoma$calendar_period <- ifelse(melanoma$year8594 == "Diagnosed 75-84", "1975–1984", "1985–1994")

# Create a survival object
surv_object <- Surv(time = melanoma$surv_mm, event = melanoma$death_cancer)

# Kaplan-Meier estimates by calendar period
km_fit <- survfit(surv_object ~ calendar_period, data = melanoma)

# Plot Kaplan-Meier curves 
plot(km_fit,
     col = c("blue", "red"),
     lwd = 2,
     conf.int = TRUE,
     main = "Kaplan-Meier Survival Curves by Calendar Period",
     xlab = "Time in months",
     ylab = "Survival Probability")
legend("topright", legend = levels(melanoma$calendar_period), 
       col = c("blue", "red"), lwd = 2)
```

Without any formal statistical test and only relying in visual comparison, we can conclude that the patient survival is superior during the time period 1985-1994 compared to 1975-1984 as the 1985-1994 survival curve is consistently higher.

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
