---
title: "Survival Analysis with Applications in Medicine: Take-home examination"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


---

## A. Weibull regression models

### Q1. For a proportional hazards model with $S_0(t)$ that is a Weibull distribution, show that survival $S(t)$ is also from a Weibull distribution.

Given a proportional hazards model with survival function  $S(t | x) = S_0(t)^{\exp(\beta x)}$ 

for time $t$ and a given covariate $x$, $S_0(t)$ is the baseline survival function, and $\beta$ is the log hazard ratio.

We have the Weibull survival function $S_0(t) = \exp(-\lambda t^{k})$ for a scale parameter $\lambda$ and a shape parameter $k$.

Substituting $S_0(t)$ in the proportional hazards model, we get the survival function as 

$S(t | x) = \left[\exp(-\lambda t^{k})\right] ^ {\exp(\beta x)}$ 

$\Rightarrow S(t | x) = \exp(-\lambda t^{k} \exp(\beta x))$

which is in form $S(t | x) = \exp(-\tilde\lambda_a t^{k})$ where $\tilde\lambda_a = \lambda \exp(\beta x)$ is the new scale parameter
for the same shape parameter $k$.

This means that the survival function $S(t|x)$ for a proportional hazards model is also from a Weibull distribution.

Here, $S(t) = E_X[S(t \mid X)]$, which will also be from a Weibull distribution.

### Q2. For an accelerated failure time model with $S_0(t)$ that is a Weibull distribution, show that survival $S(t)$ is also from a Weibull distribution.

Given an accelerated failure time model with survival function  $S(t | x) = S_0(t \exp(- \tilde\beta x))$

for time $t$ and a given covariate $x$, $S_0(t)$ is the baseline survival function, and $\tilde\beta$ is the log time ratio.

Substituting $S_0(t)$ in the accelerated failure time model, we get the survival function as

$S(t | x) = \exp(-\lambda { \left ( t \exp(- \tilde\beta x) \right ) }^{k})$

$\Rightarrow S(t | x) = \exp(-\lambda t^{k} \exp(-k \tilde\beta x))$

which is in form $S(t | x) = \exp(-\tilde\lambda_b t^{k})$ where $\tilde\lambda_b = \lambda \exp(-k \tilde\beta x)$ is the
new scale parameter for the same shape parameter $k$.

This means that the survival function $S(t|x)$ for an accelerated failure time model is also from a Weibull distribution.

Similarly, $S(t) = E_X[S(t \mid X)]$, which will also be from a Weibull distribution.


### Q3. What is the relationship between $\beta$ and $\tilde\beta$ if both models have a Weibull baseline survival function?

For the proportional hazards model, we have $S(t | x) = \exp(-\lambda t^{k} \exp(\beta x))$.

For the accelerated failure time model, we have $S(t | x) = \exp(-\lambda t^{k} \exp(-k \tilde\beta x))$.

Comparing the two models, we get $\exp(\beta x) = \exp(-k \tilde\beta x)$.

Taking the natural logarithm of both sides, we get $\beta x = -k \tilde\beta x$.

Therefore, the relationship between $\beta$ and $\tilde\beta$ is $\beta = -k \tilde\beta$ if both models have a Weibull baseline survival function.

---

## B: Interval-censored likelihood

For a data tuple $(t_i, u_i, v_i)$ where $t_i$ is the (left truncated) delayed entry time, and the event is observed in the interval $(u_i, v_i]$ for an individual $i$.

### Q1a. Express the log-likelihood in terms of Survival function $S(t)$ at time $t$:


The Likelihood for the interval-censored data: $u_i < T \leq v_i$ for an entry time $t_i$ is given by: 

$L_i = \frac{S(u_i) - S(v_i)}{S(t_i)}$

where $S(t)$ is the Survival function.

Hence, the log-likelihood in terms of Survival function $S(t)$ at time $t$ is given by $\log(S(u_i) - S(v_i)) - \log(S(t_i))$.

### Q1b. Express the log-likelihood in terms of the hazard function $h(t)$ at time $t$:

Now, for the derivation of the log-likelihood in terms of the hazard function $h(t)$ at time $t$, we need to express the Survival function $S(t)$ in terms of the hazard function $h(t)$.

We know that, the survival function $S(t)$ is given by $-log(S(t)) = H(t)$, 

where $H(t)$ is the cumulative hazard function, and $H(t) = \int_0^t h(u) du$ for the hazard function $h(t)$.

Therefore, the log-likelihood in terms of the hazard function $h(t)$ at time $t$ is:

$\log L_i = \log(S(u_i) - S(v_i)) - \log(S(t_i))$

$\Rightarrow \log L_i = \log(exp(-H(u_i)) - exp(-H(v_i))) - \log(exp(-H(t_i)))$

$\Rightarrow \log L_i = \log(exp(-H(v_i))*(exp(H(v_i) - H(u_i)) - 1)) + H(t_i)$

$\Rightarrow \log L_i = \log(exp(-H(v_i))) + \log(exp(H(v_i) - H(u_i)) - 1) + H(t_i)$

$\Rightarrow \log L_i = H(t_i) - H(v_i) + \log(exp(H(v_i) - H(u_i)) - 1)$

$\Rightarrow \log L_i = \int_0^{t_i} h(t) dt - \int_0^{v_i} h(t) dt + \log(exp(\int_{u_i}^{v_i} h(t) dt) - 1)$

$\Rightarrow \log L_i = - \int_{t_i}^{v_i} h(t) dt + \log(exp(\int_{u_i}^{v_i} h(t) dt) - 1)$

Hence, the log-likelihood in terms of the hazard function $h(t)$ at time $t$ is given by $- \int_{t_i}^{v_i} h(t) dt + \log(exp(\int_{u_i}^{v_i} h(t) dt) - 1)$.


### Q2. Can you express these data using the Surv function from the survival package? If so, show an example; if not, explain why.

Yes, we can express the interval-censored data using the `Surv` function from the `survival` package.

The `Surv` function is used to create a survival object that represents the survival time of an individual. It takes the form `Surv(time, event)` where `time` is the survival time and `event` is the event indicator.

For interval-censored data, we can use the `Surv` function as `Surv(time, time2, type = "interval2")` where `time` is the start of the interval, `time2` is the end of the interval and type `interval2` is used to indicate interval-censored data effectively.

Although, the `Surv` function doesn't support left truncation directly, we can filter out the left truncated data by taking the maximum of the entry time and the left truncation time.

Here is an example of how to express left-truncated interval-censored data using the `Surv` function:

```{r}

library(survival)

# sample data for the given data tuple structure (t_i, u_i, v_i)
d = data.frame(left_truncation_time = c(1,1,3,3,3),
               entry_time = c(0,0,2,2,2), 
               exit_time = 1:5, 
               event = c(1,0,1,0,1))

# filter out left truncated data 
d$entry_time = pmax(d$entry_time, d$left_truncation_time)

# interval-censored data
with(d, Surv(entry_time, exit_time, type="interval2")) 
```

---

## C: Truncated distributions

For a continuous random variable $T$, we have the survival function $S(t) = P(T > t)$. 

Let the quantile function $Q(p)$ such that $Q(p) = S^{-1}(1-p) \Rightarrow 1-p = S(Q(p)) = P(T > Q(p))$.

Assume that we have a truncated distribution with survival function $P(T > t| T > t_0)$ for left truncation time $t_0$.

### Q1. Express the survival function for the truncated distribution in terms of the survival function for $T$.

The survival function for the truncated distribution is given by: 

$P(T > t| T > t_0)$

$\Rightarrow \frac{P(T > t, T > t_0)}{P(T > t_0)}$ by definition of conditional probability.

$\Rightarrow \frac{P(T > t)}{P(T > t_0)}$ because $T > t, T > t_0$ implies $T > t$ as $t_0$ is the left truncation time. 

Since, $P(T > t) = S(t)$ and $P(T > t_0) = S(t_0)$ by the survival function for $T$, 

the survival function for the truncated distribution is given by $\frac{S(t)}{S(t_0)}$.


### Q2. For the truncated distribution, what is the quantile function $Q(p|t_0)$ that solves $P(T > t | T > t_0) = 1-p$ for $t$ in terms of the survival and quantile functions for $T$ at quantile (probability) $p$? 

Given $P(T > t | T > t_0) = 1-p$,

we know that $P(T > t | T > t_0) = \frac{S(t)}{S(t_0)} = 1-p$.

Therefore, the quantile function $Q(p|t_0) = t$ that solves $P(T > t | T > t_0) = 1-p$:

$S(t) = (1-p)S(t_0)$

$\Rightarrow t = S^{-1}((1-p)S(t_0))$

$\Rightarrow Q(p|t_0) = S^{-1}((1-p)S(t_0))$. 

To solve for $S^{-1}((1-p)S(t_0))$, we know that $Q(p) = S^{-1}(1-p) \rightarrow Q(1-p) = S^{-1}(p)$.

Hence, for a $y$, we can say that $S^{-1}(y) = Q(1-y)$,

$\Rightarrow Q(p|t_0) = S^{-1}((1-p)S(t_0)) =  Q(1-(1-p)S(t_0))$.

$\Rightarrow Q(p|t_0) = Q(1-(1-p)S(t_0))$.

i.e., the $p$-th quantile of left truncated distribution is the ($1 - (1-p)S(t_0)$)-th quantile of the original distribution.


### Q3. Using this algorithm, write, run and report on R code to calculate the 0.4 quantile from a truncated log-normal distribution where $T \sim LogNormal(\mu = 1, \sigma^2 = 1.2^2)$ for a log-normal distribution with mean $\mu$ and standard deviation $\sigma$ on the log scale for $t_0 = 2$.

We can derive the quantile function $Q(p= 0.4|t_0 = 2)$ as follows:

$Q(p|t_0) = Q(1 - (1-p)S(t_0))$.

To derive $S(t_0)$, we know that $S(t_0) = P(T > t_0) = 1 - P(T \leq t_0) = 1 - F(t_0)$,

where $F(t)$ is the cumulative distribution function for $T$.

For a log-normal distribution of $T$, we can compute $S(t_0)$ using `plnorm` function in R.

Then, we can substitute the $S(t_0)$ value back to $Q(1 - (1-p)S(t_0))$ and compute the quantile using `qlnorm` function in R for the probability $p = 0.4$.

```{r}
#' @param p is the probability
#' @param meanlog mean on the log scale
#' @param sdlog standard deviation on the log scale
#' #' @param t0 left truncation time
#' @return the quantile from a truncated log-normal distribution

f = function(p, meanlog, sdlog, t0) {
  S2 = 1 - plnorm(t0, meanlog, sdlog)
  Q = qlnorm(1 - (1-p)*S2, meanlog, sdlog)
  return(Q)
}

# function call
f(p = 0.4, meanlog = 1, sdlog = 1.2, t0 = 2)

```

Hence, the 0.4 quantile from a truncated log-normal distribution with $T \sim LogNormal(\mu = 1, \sigma^2 = 1.2^2)$ for $t_0 = 2$ is approximately 4.172.

### Q4. Check your value of the 0.4 quantile from the truncated log-normal distribution in question C3 by given random sampling code. 

Given `R` code to return a vector of random numbers $t$ sampled from a truncated log-normal distribution:

```{r}
#' @param n the number of random numbers
#' @param meanlog mean on the log scale
#' @param sdlog sd on the log scale
#' @param t0 left truncation time(s)
#' @return vector of random numbers drawn from a truncated log-normal distribution
rtrunc_lnorm = function(n, meanlog, sdlog, t0) {
    y = rlnorm(n, meanlog, sdlog)
    while (any(y<t0))
        y[y<t0] = rlnorm(n, meanlog, sdlog)[y<t0]
    y
}
```

Now, let's try to compute the 0.4 quantile for the above random numbers using the `quantile` function in R for large sample sizes of $n = 10,000$ and $n = 100,000$.

```{r}
# set seed for reproducibility
set.seed(123)

# generate 10,000 random numbers from truncated log-normal distribution
y = rtrunc_lnorm(n = 10000, meanlog = 1, sdlog = 1.2, t0 = 2)

# calculate the 0.4 quantile
quantile(y, 0.4)

# generate 100,000 random numbers from truncated log-normal distribution
y = rtrunc_lnorm(n = 100000, meanlog = 1, sdlog = 1.2, t0 = 2)

# calculate the 0.4 quantile
quantile(y, 0.4)

```

We can see that the 0.4 quantile is approximately 4.149 and 4.165 by randomly sampling $t$ values from a truncated log-normal distribution of $T$ for $n = 10,000$ and $n = 100,000$ respectively. This is very close to the value of 4.172 obtained from the algorithm in question C3.


---

## D: Cox’s partial likelihood with a time-varying effects 

### Q1. Let the right censored data tuple $(t_i, \delta_i, x_i(.))$ for individual $i \in \{1, 2, ..., n\}$, with follow-up from time 0 to time $t_i$, event indicator $\delta_i$ (with value 1 if the event is observed at time $t_i$, otherwise censored with value 0), and time-varying effects $x_i(t)$. Let the time-varying hazard ratio be $exp(x_i(t)^{T} \beta)$ for regression coefficients $\beta$. Let the risk set $R(t_i)$ be the set of individuals $\{j : t_j \geq t_i\}$. Assume that there are no tied event times. Write out the partial likelihood $L(\beta)$. 

The cox proportional hazards model has the hazard at time $t$ given time-varying covariates $x(t)$ as 

$h(t|x) = exp(x(t)^{T} \beta) h_0(t)$, where $h_0(t)$ is the baseline hazard function.

The partial likelihood function $L(\beta)$ is the product over all individuals who experience an event ($\delta_i = 1$) of the conditional probability that that particular individual experiences the event at time $t_i$, given that the event occurs at that time within the risk set $R(t_i)$.

For an individual $i$ who experiences an event at time $t_i$, the conditional probability is given by the ratio of their hazard to the sum of the hazards of all individuals in the risk set at time $t_i$:

$\frac{h_i(t_i|x_i)}{\sum_{j \in R(t_i)} h_j(t_i|x_i)}$

Substituting the above form of the hazard function, the baseline hazard $h_0(t_i)$ cancels out from the numerator and the denominator:

$\frac{exp(x_i(t_i)^{T} \beta) h_0(t_i)}{\sum_{j \in R(t_i)} exp(x_j(t_i)^{T} \beta) h_0(t_i)} = \frac{exp(x_i(t_i)^{T} \beta)}{\sum_{j \in R(t_i)} exp(x_j(t_i)^{T} \beta)}$.

Hence, the partial likelihood function $L(\beta)$ is the product of these conditional probabilities over all individuals for whom an event is observed: 

$L(\beta) = \prod_{i=1}^{n} \left ( \frac{exp(x_i(t_i)^{T} \beta)}{\sum_{j \in R(t_i)} exp(x_j(t_i)^{T} \beta)} \right)^{\delta_i}$

Note that the $x_j$ in the denominator are evaluated at time $t_i$ .

### Q2. Analytically derive the gradient (or score) $\frac{d\log(L)}{d\beta_k}$.

The log partial likelihood function is given by:

$\log(L(\beta)) = \sum_{i=1}^{n} \delta_i \left ( x_i(t_i)^{T} \beta - \log \left ( \sum_{j \in R(t_i)} exp(x_j(t_i)^{T} \beta) \right ) \right)$.

The gradient of the log partial likelihood function with respect to the $k$-th element of $\beta$ is:

$\frac{d\log(L)}{d\beta_k} = \sum_{i=1}^{n} \delta_i \left ( x_{ik}(t_i) - \frac{\sum_{j \in R(t_i)} x_{jk}(t_i) exp(x_j(t_i)^{T} \beta)}{\sum_{j \in R(t_i)} exp(x_j(t_i)^{T} \beta)} \right)$.

where $x_{ik}(t_i)$ is the $k$-th element of the time-varying effects vector $x_i(t_i)$ for individual $i$ and 

$x_{jk}(t_i)$ is the $k$-th element of the time-varying effects vector $x_j(t_i)$ for individual $j$ in the risk set $R(t_i)$ at time $t_i$.


### Q3. Let a binary exposure be defined by $z_i$ for individual $i$ and let $x_i(t) = (z_i, z_i t)^T$. Write out a formula for the hazard ratio as a function of time $t$ for those exposed compared with those not exposed.

Considering an exposure $z_i$, the hazard function for an individual $i$ is given by:

$h_i(t|x_i) = exp(x_i(t)^{T} \beta) h_0(t)$.

$\Rightarrow exp(z_i \beta_1 + z_i t \beta_2) h_0(t)$.

The hazard ratio at time $t$ for those exposed ($z_i = 1$) compared with those not exposed ($z_i = 0$) is given by:

$\frac{h_i(t|x_i = (1, t)^T)}{h_i(t|x_i = (0, 0)^T)} = \frac{exp(\beta_1 + t \beta_2) h_0(t)}{exp(0) h_0(t)} = exp(\beta_1 + t \beta_2)$.

### Q4. The following code is used to investigate whether the hazard ratio between distant and localised cancer varies by time. Write out the regression model and carefully interpret the four parameters. 

Given the following code:

```{r warning=FALSE, message=FALSE}
library(survival)
library(biostat3)
transform(biostat3::colon, stage=relevel(stage,"Localised")) |>
    coxph(Surv(surv_mm,status=="Dead: cancer")~stage+tt(stage), data=_,
          tt=function(x,t,...) (x=="Distant")*t/12) |>
    summary()
```

From the above code, the cox regression model being fitted can be written as: 

$\mathbf{h(t| stage) = h_0(t) exp(\beta_1 \times stageUnknown + \beta_2 \times stageRegional + \beta_3 \times stageDistant + \beta_4 \times tt(stage))}$.

where $h(t| stage)$ is the hazard function at time $t$ given the stage of cancer, $h_0(t)$ is the baseline hazard function, and $\beta_1, \beta_2, \beta_3, \beta_4$ are the regression coefficients for the stage of cancer.

The model includes three stages of cancer, such as Unknown, Regional, Distant, and a time-varying effect `tt(stage)`, defined as $(x == Distant) \times t/12$, which is only applicable when the stage is Distant. 

Here, Localised is the reference level for the stage of cancer.

The four parameters in question are:

 - $\beta_1$: The log hazard ratio between the Unknown stage and the Localised stage of cancer. 
 - $\beta_2$: The log hazard ratio between the Regional stage and the Localised stage of cancer.
 - $\beta_3$: The log hazard ratio between the Distant stage and the Localised stage of cancer at $t=0$.
 - $\beta_4$: The time-dependent change in the log hazard ratio between the Distant stage and the Localised stage of cancer per year. 
 
The p-value of each parameter is a very small value, much less than 0.05, indicating that the parameters are all statistically significant. This means that all stages of cancer have a significant impact on the hazard of death from cancer compared to the Localised stage. 

The time-varying effect for the Distant stage is also significant, indicating that the hazard ratio between Distant and Localised cancer varies over time.


Let us interpret these parameters using the hazard ratios mentioned in the summary:

 - Patients with Unknown stage cancer have $exp(\beta_1) \sim$ **2.56** times higher risk of death from cancer compared to patients with Localised stage cancer.
 - Patients with Regional stage cancer have $exp(\beta_2) \sim$ **2.23** times higher risk of death from cancer compared to patients with Localised stage cancer.
 - Patients with Distant stage cancer have $exp(\beta_3) \sim$ **9.20** times higher risk of death from cancer compared to patients with Localised stage cancer.
 - The hazard for Distant stage cancer decreases over time at a rate of $1 - exp(\beta_4) \sim 1- 0.88 = 0.12$, meaning approximately a **12%** reduction in hazard per year.
 
The 95% confidence interval for the hazard ratio of time varying effect is (0.86, 0.91), which indicates that the hazard for Distant stage cancer decreases by approximately 9% to 14% per year compared to Localised stage cancer with 95% confidence.




---

## E: Data analysis of a randomised controlled trial for hormonal treatment of breast cancer patients in Germany

---

## F: Analysis plan for a randomised controlled trial

---

I am declaring that I have used generative artificial intelligence (GAI) to assist me in completing this assignment in the form of Github co-pilot to auto-complete my explanations.


This assignment took me approximately 10 hours to complete.

